\bf Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-07-05 paper.qmd
finished conclusion

---
title: "Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
author: 
  - name: Annalisa Calvi
    affiliations:
      name: Monash University
      department: School of Mathematics
  - name: Ursula Laa
    affiliations:
      name: University of Natural Resources and Life Sciences, Vienna
      department: Institute of Statistics
    email: ursula.laa@boku.ac.at
    ORCID: 0000-0002-0249-6439
  - name: Dianne Cook
    affiliations:
      name: Monash University
      department: Department of Econometrics and Business Statistics
    email: dicook@monash.edu
    ORCID: 0000-0002-3813-7155
format:
  jasa-pdf:
    keep-tex: true    
abstract: |
  Many data problems contain some reference or normal conditions, upon which to compare newly collected data. This scenario occurs in data collected as part of clinical trials to detect adverse events, and or for measuring climate change against historical norms. The data is typically multivariate, and often the normal ranges are specified by a multivariate normal distribution. The work presented in this paper develops methods to compare the new sample against the reference distribution with high-dimensional visualisation. It uses a projection pursuit guided tour to produce a sequence of low-dimensional projections steered towards those where the new sample is most different from the reference. A new projection pursuit index is defined, and the drawing of the projected ellipse is computed analytically. The methods are implemented in the R package, `tourr`. 
keywords:
  - exploratory data analysis
  - statistical graphics
  - dimension reduction
  - high-dimensional visualisation
  - climate change
  - clinical trials
bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{amsthm}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newtheorem*{theorem}{Theorem}
---

@@

Say something about the relationship with @directional-outlyingness and Stahel-Donoho outlyingness. This method could be implemented in the way we used to show ellipses, generate points on the surface of the irregular shape, overlay the data on this and make projections.

Difference from general outlier detection, see use of Mahalanobis distance in @filzmoser2018.

Potential new directions.

>>>>

---
title: "Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
author: 
  - name: Annalisa Calvi
    affiliations:
      name: Monash University
      department: School of Mathematics
  - name: Ursula Laa
    affiliations:
      name: University of Natural Resources and Life Sciences, Vienna
      department: Institute of Statistics
    email: ursula.laa@boku.ac.at
    ORCID: 0000-0002-0249-6439
  - name: Dianne Cook
    affiliations:
      name: Monash University
      department: Department of Econometrics and Business Statistics
    email: dicook@monash.edu
    ORCID: 0000-0002-3813-7155
format:
  jasa-pdf:
    keep-tex: true    
abstract: |
  Many data problems contain some reference or normal conditions, upon which to compare newly collected data. This scenario occurs in data collected as part of clinical trials to detect adverse events, and or for measuring climate change against historical norms. The data is typically multivariate, and often the normal ranges are specified by a multivariate normal distribution. The work presented in this paper develops methods to compare the new sample against the reference distribution with high-dimensional visualisation. It uses a projection pursuit guided tour to produce a sequence of low-dimensional projections steered towards those where the new sample is most different from the reference. A new projection pursuit index is defined, and the drawing of the projected ellipse is computed analytically. The methods are implemented in the R package, `tourr`. 
keywords:
  - exploratory data analysis
  - statistical graphics
  - dimension reduction
  - high-dimensional visualisation
  - climate change
  - clinical trials
bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{amsthm}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newcommand\bD{$1\text{-}D$}
  \newtheorem*{theorem}{Theorem}
---

@@

This paper has provided a new projection pursuit index for comparing a sample of data against a multivariate normal reference distribution. It has also provided an analytical result for drawing \gD{} projections of the \pD{} ellipse corrsponding to the normal reference distribution. These combined provide new ways to visualise multivariate data, for this particular scenario.

The work is related to outlier detection methods [see, @filzmoser2018, for example], particularly those that use robust statistics, as illustrated in the weather example. Methods discussed in @directional-outlyingness and @donoho1982 describe detecting outlyingness using projections of the data. When these projections are collected and used as a large set, they define a high-dimensional confidence region of an indeterminate shape. There is no analytical definition. However, we could explore the data relative to the high-dimensional shapes using the same approach described in @sec-background. You would generate points on the surface of the irregular shape, overlay the data on this to visualise with a tour.

The methods on directional outlyingness or Stahel-Donoho outlyingness, described above, leng themselves to new potential \bD{} projection pursuit indexes. These would integrate nicely into a \bD{} projection pursuit guided tour to interactively visualise the potential outliers relative to the rest of the sample.

Another potential direction of this work is with model-based clustering using Gaussian mixtures [@FR02], where ellipses form the basis of the model. Currently, the `mclust` software [@mclust] has the capacity to show \gD{} ellipses for two variables, or axis parallel \gD{} ellipses for \pD{}. The equations developed here for drawing the \gD{} projection of a \pD{} ellipse would provide more versatile visualisation for examining the model-based clustering fits.

### 2024-07-05 paper.tex
finished conclusion

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$}
\newcommand\kD{$k\text{-}D$}
\newcommand\dD{$d\text{-}D$}
\newcommand\gD{$2\text{-}D$}
\newtheorem*{theorem}{Theorem}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

Say something about the relationship with
\citet{directional-outlyingness} and Stahel-Donoho outlyingness. This
method could be implemented in the way we used to show ellipses,
generate points on the surface of the irregular shape, overlay the data
on this and make projections.

Difference from general outlier detection, see use of Mahalanobis
distance in \citet{filzmoser2018}.

Potential new directions.

>>>>

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$}
\newcommand\kD{$k\text{-}D$}
\newcommand\dD{$d\text{-}D$}
\newcommand\gD{$2\text{-}D$}
\newcommand\bD{$1\text{-}D$}
\newtheorem*{theorem}{Theorem}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

This paper has provided a new projection pursuit index for comparing a
sample of data against a multivariate normal reference distribution. It
has also provided an analytical result for drawing \gD{} projections of
the \pD{} ellipse corrsponding to the normal reference distribution.
These combined provide new ways to visualise multivariate data, for this
particular scenario.

The work is related to outlier detection methods \citep[see,][for
example]{filzmoser2018}, particularly those that use robust statistics,
as illustrated in the weather example. Methods discussed in
\citet{directional-outlyingness} and \citet{donoho1982} describe
detecting outlyingness using projections of the data. When these
projections are collected and used as a large set, they define a
high-dimensional confidence region of an indeterminate shape. There is
no analytical definition. However, we could explore the data relative to
the high-dimensional shapes using the same approach described in
Section~\ref{sec-background}. You would generate points on the surface
of the irregular shape, overlay the data on this to visualise with a
tour.

The methods on directional outlyingness or Stahel-Donoho outlyingness,
described above, leng themselves to new potential \bD{} projection
pursuit indexes. These would integrate nicely into a \bD{} projection
pursuit guided tour to interactively visualise the potential outliers
relative to the rest of the sample.

Another potential direction of this work is with model-based clustering
using Gaussian mixtures \citep{FR02}, where ellipses form the basis of
the model. Currently, the \texttt{mclust} software \citep{mclust} has
the capacity to show \gD{} ellipses for two variables, or axis parallel
\gD{} ellipses for \pD{}. The equations developed here for drawing the
\gD{} projection of a \pD{} ellipse would provide more versatile
visualisation for examining the model-based clustering fits.
